---
title: "Lab 5 MATH 342W"
author: "Bryan Nevarez"
output: pdf_document
date: "11:59PM March 12"
---

Write a function spec'd as follows:

```{r}
#' Orthogonal Projection
#'
#' Projects vector a onto v.
#'
#' @param a   the vector to project
#' @param v   the vector projected onto
#'
#' @returns   a list of two vectors, the orthogonal projection parallel to v named a_parallel, 
#'            and the orthogonal error orthogonal to v called a_perpendicular
orthogonal_projection = function(a, v){
  a_parallel = (t(t(v)) %*% t(v) / (sum(v^2))) %*% t(t(a))
  list(a_parallel = a_parallel, a_perpendicular = t(t(a)) - a_parallel)
}
```

Provide predictions for each of these computations and then run them to make sure you're correct.

```{r}
orthogonal_projection(c(1,2,3,4), c(1,2,3,4))
#prediction: 
orthogonal_projection(c(1, 2, 3, 4), c(0, 2, 0, -1))
#prediction:
result = orthogonal_projection(c(2, 6, 7, 3), c(1, 3, 5, 7))
t(result$a_parallel) %*% result$a_perpendicular
#prediction:
result$a_parallel + result$a_perpendicular
#prediction:
result$a_parallel / c(1, 3, 5, 7)
#prediction:
```



Create a vector y by simulating n = 100 standard iid normals. Create a matrix of size 100 x 2 and populate the first column by all ones (for the intercept) and the second column by 100 standard iid normals. Find the R^2 of an OLS regression of `y ~ X`. Use matrix algebra.

```{r}
n = 100
y = rnorm(n) # 100 standard iid
X = cbind(1, rnorm(n))

yhat = X %*% solve(t(X) %*% X) %*% t(X) %*% y 
ybar = mean(y)

SSR = sum((yhat - ybar)^2)
SST = sum((y - ybar)^2)

RSQ = SSR / SST

```

Write a for loop to each time bind a new column of 100 standard iid normals to the matrix X and find the R^2 each time until the number of columns is 100. Create a vector to save all R^2's. What happened??

```{r}
rsqs = array(NA, n)
rsqs[2] = RSQ
for(j in 3:n){
  X = cbind(X,rnorm(n))
  yhat = X %*% solve(t(X) %*% X) %*% t(X) %*% y 
  ybar = mean(y)

  SSR = sum((yhat - ybar)^2)
  SST = sum((y - ybar)^2)

  rsqs[j] = SSR / SST
}
print(rsqs)


```

Test that the projection matrix onto this X is the same as I_n. You may have to vectorize the matrices in the `expect_equal` function for the test to work.

```{r}
pacman::p_load(testthat)
expect_equal( X %*% solve(t(X) %*% X) %*% t(X), diag(n))
```

Add one final column to X to bring the number of columns to 101. Then try to compute R^2. What happens? 

There is an error when we try to find the inverse of t(X) %*% X due to it now being a singular matrix.

```{r}
X = cbind(X, rnorm(n))
yhat = X %*% solve(t(X) %*% X) %*% t(X) %*% y
```

Why does this make sense?

Yes, because the X is now of rank n yet has n+1 columns. So when you try to find the H hat matrix the computation of t(X) %*% X is now a matrix of rank n yet of size (n+1) by (n+1) and thus this matrix is singular and therefore our R^2 will "blow up."

Let's use the Boston Housing Data for the following exercises

```{r}
y = MASS::Boston$medv
X = model.matrix(medv ~ ., MASS::Boston)
p_plus_one = ncol(X)
n = nrow(X)
```

Using your function `orthogonal_projection` orthogonally project onto the column space of X by projecting y on each vector of X individually and adding up the projections and call the sum `yhat_naive`.

```{r}
yhat_naive = matrix(0, nrow = n, ncol = 1)
for(j in 1:p_plus_one){
  yhat_naive = yhat_naive + orthogonal_projection(y, X[, j])$a_parallel
}
```

How much double counting occurred? Measure the magnitude relative to the true LS orthogonal projection.

```{r}
yhat = lm(medv ~ ., MASS::Boston)$fitted.values
sqrt(sum(yhat_naive^2)) / sqrt(sum(yhat^2))
```

Is this ratio expected? Why or why not?

Yes, it is expected because we are projecting onto a column space of X where the columns themselves are not necessarily orthogonal to one another. 

Convert X into V where V has the same column space as X but has orthogonal columns. You can use the function `orthogonal_projection`. This is the Gram-Schmidt orthogonalization algorithm (part A).

```{r}
V = matrix(NA, nrow = n, ncol = p_plus_one)
V[, 1] = X[, 1]
for (j in 2 : ncol(X)){
  V[, j] = X[, j]
  
  for (k in 1:(j-1)){
    v_k = V[, k]
    V[, j] = V[, j, drop = FALSE] - orthogonal_projection(X[, j], v_k)$a_parallel
  }
}
```

Convert V into Q whose columns are the same except normalized. This is the Gram-Schmidt orthogonalization algorithm (part B).

```{r}
Q = matrix(NA, nrow = n, ncol = p_plus_one)
for(j in 1 : p_plus_one){
  Q[, j] = V[, j] / sqrt(sum(V[, j]^2))
}
rm(V)
```

Verify Q^T Q is I_{p+1} i.e. Q is an orthonormal matrix.

```{r}
expect_equal(t(Q) %*% Q, diag(p_plus_one))
```

Is your Q the same as what results from R's built-in QR-decomposition function?

No, my Q is not the same from R's built-in QR-decomposition function. 

```{r}
Q_from_Rs_builtin = qr.Q(qr(X))
expect_equal(Q,Q_from_Rs_builtin)
```
 
Is this expected? Why did this happen?

Yes, this is expected because the Q matrix is not a unique matrix due to the fact that there are multiple bases that give us the same span being the columnspace of X. 

Project y onto colsp[Q] and verify it is the same as the OLS fit. You may have to use the function `unname` to compare the vectors since they the entries will likely have different names.

```{r}
yhat_Q = Q %*% t(Q) %*% y
yhat_from_R = Q_from_Rs_builtin %*% t(Q_from_Rs_builtin) %*% y
expect_equal(yhat_Q,yhat_from_R)
expect_equal(yhat_Q, yhat)
```

Project y onto colsp[Q] one by one and verify it sums to be the projection onto the whole space.

```{r}
yhat_naive = y_hat_Q = Q %*% t(Q) %*% y
yhat_naive = matrix(0, nrow = n, ncol = 1)
for(j in 1: ncol(Q)){
  yhat_naive = yhat_naive + orthogonal_projection(y, Q[, j])$a_parallel
}
expect_equal(y_hat_Q, yhat_naive)
length(yhat_naive)
cbind(y_hat_Q, yhat_naive)
# y_hat_Q and yhat_naive are the same!
```

Split the Boston Housing Data into a training set and a test set where the training set is 80% of the observations. Do so at random.

```{r}
K = 5
n_test = round(n * 1 / K)
n_train = n - n_test
train_indices = sample(1:n, n_train, replace = FALSE)
X_train = X[train_indices, ]
y_train = y[train_indices]
test_indices = setdiff(1:n, train_indices)
X_test = X[test_indices, ]
y_test = y[test_indices]
```

Fit an OLS model. Find the s_e in sample and out of sample. Which one is greater? Note: we are now using s_e and not RMSE since RMSE has the n-(p + 1) in the denominator not n-1 which attempts to de-bias the error estimate by inflating the estimate when overfitting in high p. Again, we're just using `sd(e)`, the sample standard deviation of the residuals.

```{r}
b = solve(t(X_train) %*% X_train) %*% t(X_train) %*% y_train
yhat_train = X_train %*% b
e_train = y_train - yhat_train
rmse_train = sd(e_train)
yhat_test = X_test %*% b
e_test = y_test - yhat_test
rmse_test = sd(e_test) 
rmse_train
rmse_test
```

Do these two exercises `Nsim = 1000` times and find the average difference between s_e and ooss_e. 

```{r}
Nsim = 1000
rmse_trains = array(NA, Nsim)
rmse_tests = array(NA, Nsim)
for(nsim in 1:Nsim){
  train_indices = sample(1:n, n_train, replace = FALSE)
  X_train = X[train_indices, ]
  y_train = y[train_indices]
  test_indices = setdiff(1:n, train_indices)
  X_test = X[test_indices, ]
  y_test = y[test_indices]
  
  b = solve(t(X_train) %*% X_train) %*% t(X_train) %*% y_train
  yhat_train = X_train %*% b
  e_train = y_train - yhat_train
  rmse_trains[nsim] = sd(e_train)
  yhat_test = X_test %*% b
  e_test = y_test - yhat_test
  rmse_tests[nsim] = sd(e_test) 
}
mean(rmse_trains)
mean(rmse_tests)
```

We'll now add random junk to the data so that `p_plus_one = n_train` and create a new data matrix `X_with_junk.`

```{r}
X_with_junk = cbind(X, matrix(rnorm(n * (n_train - p_plus_one)), nrow = n))
dim(X)
dim(X_with_junk)
```

Repeat the exercise above measuring the average s_e and ooss_e but this time record these metrics by number of features used. That is, do it for the first column of `X_with_junk` (the intercept column), then do it for the first and second columns, then the first three columns, etc until you do it for all columns of `X_with_junk`. Save these in `s_e_by_p` and `ooss_e_by_p`.

```{r}
s_e_by_p = array(NA, n_train)
ooss_e_by_p = array(NA, n_train)

X_with_junk_train = X_with_junk[train_indices, ]
y_train = y[train_indices]
X_with_junk_test = X_with_junk[test_indices, ]
y_test = y[test_indices]

for(j in 1:n_train){
  X_with_junk_train_j = X_with_junk_train[, 1:j, drop = FALSE]
  b = solve(t(X_with_junk_train_j) %*% X_with_junk_train_j) %*% t(X_with_junk_train_j) %*% y_train
  yhat_train = X_with_junk_train_j %*% b
  e_train = y_train - yhat_train
  s_e_by_p[j] = sd(e_train)
  yhat_test = X_with_junk_test[, 1:j, drop = FALSE] %*% b
  e_test = y_test - yhat_test
  ooss_e_by_p[j] = sd(e_test)
}
```

You can graph them here:

```{r}
pacman::p_load(ggplot2)
ggplot(
  rbind(
    data.frame(s_e = s_e_by_p, p = 1 : n_train, series = "in-sample"),
    data.frame(s_e = ooss_e_by_p, p = 1 : n_train, series = "out-of-sample")
  )) +
  geom_line(aes(x = p, y = s_e, col = series)) 
```
 
Is this shape expected? Explain.

Yes, this shape is expected since the red line which represents the standard error represents an eventual "overfitting" of the data and we see how the error goes to 0 (R^2 goes to 1). However, the additional augmenting of features, i.e., columns to the design X matrix does starts to precipitously decrease the performance of how well the model does its prediction with out-of-sample data as seen by the blue line. 

Now repeat the exercise above except use 5-fold CV (K=5 cross validation) for each p. The code below will also plot the oos RMSE. This oos RMSE curve should be similar to the curve in the above problem, but now it will be more stable. 


```{r}
K = 5
# oos_e_by_p_k = matrix(NA, nrow = n, ncol = n) #save all residuals here - each row are the residuals for number of features = j
set.seed(1984)
temp = rnorm(n)
folds_vec = cut(temp, breaks = quantile(temp, seq(0, 1, length.out = K + 1)), include.lowest = TRUE, labels = FALSE)
rm(temp)
head(folds_vec, 100)
min_n_train = min(n - table(folds_vec))
ooss_e_by_p = array(NA, n_train)
oos_se_by_p_by_k = matrix(NA, nrow = min_n_train, ncol = K)
for (j in 1:min_n_train){
  all_e_test = array(NA, n)
  for(k in 1:K){
    test_indices = which(folds_vec == k)
    train_indices = setdiff(1:n, test_indices)
    X_with_junk_train = X_with_junk[train_indices, ]
    y_train = y[train_indices]
    X_with_junk_test = X_with_junk[test_indices, ]
    y_test = y[test_indices]
    X_with_junk_train_j = X_with_junk_train[ , 1:j, drop = FALSE]
    b = solve(t(X_with_junk_train_j) %*% X_with_junk_train_j) %*% t(X_with_junk_train_j) %*% y_train
    yhat_test = X_with_junk_test[ , 1:j, drop = FALSE] %*% b
    e_k = y_test - yhat_test
    all_e_test[test_indices] = e_k
    oos_se_by_p_by_k[j , k] = sd(e_k)
  }
  ooss_e_by_p[j] = sd(all_e_test)
}
```



```{r}
#now plot it
pacman::p_load(ggplot2)
ggplot(data.frame(
    s_e = ooss_e_by_p,
    p = 1 : n_train
  )) +
  geom_line(aes(x = p, y = s_e))
```


Even though the concept of confidence intervals (CIs) will not be on the midterm, construct 95% CIs for each of the oosRMSE measurements by number of features, p. A CI is a real-number interval with a lower bound and upper bound. The formula for the CI is [s_e - 2 * s_s_e, s_e + 2 * s_s_e].


```{r}
s_s_es_by_j = apply(oos_se_by_p_by_k, 1, sd)
cbind(
  ooss_e_by_p - 2 * s_s_es_by_j /sqrt(K), 
  ooss_e_by_p + 2 * s_s_es_by_j / sqrt(K)
)
  
```


