---
title: "Final Project for Math 642"
author: "Bryan Nevarez"
date: "Sunday, May 26th, 2024"
output: pdf_document
---

# Packages/Libraries Setup
```{r,set.seed(342)}
# remove everything from the current workspace
rm(list = ls())
#Set cache for seed
knitr::opts_chunk$set(cache = T)
#Memory allocation for Java ~10gb and Garbage Collection
options(java.parameters = c("-XX:+UseConcMarkSweepGC", "-Xmx10000m"))
#Packages to load
pacman::p_load(
  ggplot2,
  tidyverse,
  data.table,
  R.utils,
  magrittr,
  dplyr,
  testthat,
  YARF,
  lubridate,
  missForest,
  parallel,
  doParallel,
  caret,
  glmnet,
  mlr,
  mlr3
)
#Set CPU cores for YARF
num_of_cores = 8
set_YARF_num_cores(num_of_cores)
#Initialize rJava
library(rJava)
gc()
.jinit()
```

# Data Import
```{r}
library(readr)
#Read in the given raw housing data file 
raw_housing_data = read_csv("housing_data_2016_2017.csv")

# Read in the nyc data per zip code found from the internet
nyc_data_per_zipcode = read_csv("New York City_CensusZipCodeTabulationArea.csv")

# checks the specifications of the columns for both raw data sets
spec(raw_housing_data)
spec(nyc_data_per_zipcode)

# View information on both raw data sets
skimr::skim(raw_housing_data)
skimr::skim(nyc_data_per_zipcode)
```

# Data Preparation, Pt. I: raw_housing_data 
```{r}
# make raw_housing_data into a tibble
housing_data_tbl = as_tibble(raw_housing_data)

# drop the first 28 columns of housing_data_tbl 
housing_data_tbl %<>%
  select(-c(1:28))

# drop the model_type column (and the url column) due to it collinearity with other columns
housing_data_tbl %<>%
  select(-model_type, -url)

# inside view to housing_data_tbl
skimr::skim(housing_data_tbl)

# remove raw_housing_data
rm(raw_housing_data)
```

# Data Prepartion, Pt. II: nyc_data_per_zipcode 
```{r}
# Make nyc zip code dataset into a tibble
nyc_data_per_zipcode_tbl = as_tibble(nyc_data_per_zipcode)

# deleting all rows deemed to be irrelevant or uninformative
nyc_data_per_zipcode_tbl %<>%
  select(-c('placeDcid','Date:Median_Age_Person', 'Source:Median_Age_Person','Date:Median_Income_Person','Source:Median_Income_Person','Date:Count_Person','Source:Count_Person','Date:Count_HousingUnit','Source:Count_HousingUnit','Date:Monthly_Median_GrossRent_HousingUnit','Value:Monthly_Median_GrossRent_HousingUnit','Source:Monthly_Median_GrossRent_HousingUnit'))

# renaming the columns 
nyc_data_per_zipcode_tbl %<>%
  rename(zip_code = placeName,  median_age = 'Value:Median_Age_Person', median_income = 'Value:Median_Income_Person' , population = 'Value:Count_Person', housing_units = 'Value:Count_HousingUnit')

# inside view of nyc_data_per_zipcode_tbl
skimr::skim(nyc_data_per_zipcode_tbl)

# remove nyc_data_per_zipcode
rm(nyc_data_per_zipcode)
```

# Data Cleaning I (Errors, Duplicates, Outliers Handling)
```{r}
# remove the dollar symbol from each of the entries in the relevant columns 
#housing_data_tbl$common_charges = gsub("\\$", "", housing_data_tbl$common_charges)
#housing_data_tbl$maintenance_cost = gsub("\\$", "", housing_data_tbl$maintenance_cost)
#housing_data_tbl$parking_charges = gsub("\\$", "", housing_data_tbl$parking_charges)
#housing_data_tbl$total_taxes = gsub("\\$", "", housing_data_tbl$total_taxes)
#housing_data_tbl$sale_price = gsub("\\$", "", housing_data_tbl$sale_price)
#housing_data_tbl$sale_price = gsub("\\,", "", housing_data_tbl$sale_price)
#housing_data_tbl$listing_price_to_nearest_1000 = gsub("\\$", "", #housing_data_tbl$listing_price_to_nearest_1000)

# one line piping commmand to get rid of dollar sign and commas from the above commented cols
housing_data_tbl %<>%
  mutate(across(c(3,13,19,21,23,25), gsub, pattern = "\\$|,", replacement = ""))

# cast various features into appropriate data type
housing_data_tbl %<>% 
  mutate(common_charges = as.numeric(common_charges),
         maintenance_cost = as.numeric(maintenance_cost),
         parking_charges = as.numeric(parking_charges),
         total_taxes = as.numeric(total_taxes), 
         listing_price_to_nearest_1000 = as.numeric(listing_price_to_nearest_1000),
         sale_price = as.numeric(sale_price),
         approx_year_built = as.integer(housing_data_tbl$approx_year_built))


# clear out human errors in character-type features

# str_extract() extracts the first occurrence of a pattern matching a five-digit zip code (\\b\\d{5}\\b) from each element of the full_address_or_zip_code column.
# The result is stored in a new column called zip_code in housing_data_tbl.
housing_data_tbl = housing_data_tbl %<>%
  mutate(zip_code = sub(".*\\b(\\d{5})\\b.*", "\\1", housing_data_tbl$full_address_or_zip_code)) 

# drop full_address_or_zip_code column and date of sale after extraction operation 
housing_data_tbl %<>%
  select(-full_address_or_zip_code, -date_of_sale )

# the reasonably small number of misspellings,inaccuracies in the certain data entries 
# have been carefully corrected and standardized one-by-one in the code below

# Zip code Column
housing_data_tbl %<>%
  mutate(zip_code = replace(zip_code, zip_code == "138-35 Elder Ave,  Flushing NY, 1135", "11355")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "32-42 89th St,  E. Elmhurst NY, 1136", "11369")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "34-30 78th St,  Jackson Heights NY, 1137", "11375")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "35-25 77 St,  Jackson Heights NY, 1137", "11375")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "61-20 Grand Central Pky,  Forest Hills NY, 11375", "11355")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "78-07 Springfield Blvd,  Bayside NY, 1136", "11364")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "80-35 Springfield Blvd,  Queens Village NY, 11429", "11355")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "138-35 Elder Ave,  Flushing NY, 1135", "11355")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "32-42 89th St,  E. Elmhurst NY, 1136", "11369")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "35-25 77 St,  Jackson Heights NY, 1137", "11372")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "61-20 Grand Central Pky,  Forest Hills NY, 11375", "11355")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "78-07 Springfield Blvd,  Bayside NY, 1136", "11364")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "80-35 Springfield Blvd,  Queens Village NY, 11429", "11355")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "61-20 Grand Central Pky,  Forest Hills NY, 1137", "11375")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "80-35 Springfield Blvd,  Queens Village NY, 1142", "11429")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "61-20 Grand Central Pky,  Forest Hills NY, 1137", "11375")) %>%
  mutate(zip_code = replace(zip_code, zip_code == "80-35 Springfield Blvd,  Queens Village NY, 1142", "11429")) 

# cats_allowed column
housing_data_tbl %<>%
  mutate(cats_allowed = replace(cats_allowed, cats_allowed == "y", "yes"))

# dogs_allowed column
housing_data_tbl %<>%
  mutate(dogs_allowed = replace(dogs_allowed, dogs_allowed == "yes89", "yes"))

# dining_room_type column
housing_data_tbl %<>%
  mutate(dining_room_type = replace(dining_room_type, dining_room_type == "yes89", "yes"))

# fuel_type column
housing_data_tbl %<>%
  mutate(fuel_type = replace(fuel_type, fuel_type == "Other", "other")) %>%
  mutate(fuel_type = replace(fuel_type, fuel_type == "none", NA))

# kitchen_type column
housing_data_tbl %<>%
  mutate(kitchen_type = replace(kitchen_type, kitchen_type == "eat in", "eat-in")) %>%
  mutate(kitchen_type = replace(kitchen_type, kitchen_type == "Eat In", "eat-in")) %>%
  mutate(kitchen_type = replace(kitchen_type, kitchen_type == "Eat in", "eat-in")) %>%
  mutate(kitchen_type = replace(kitchen_type, kitchen_type == "eatin", "eat-in")) %>%
  mutate(kitchen_type = replace(kitchen_type, kitchen_type == "efficiency kitchene", "efficiency")) %>%
  mutate(kitchen_type = replace(kitchen_type, kitchen_type == "efficiency kitchen", "efficiency")) %>%
  mutate(kitchen_type = replace(kitchen_type, kitchen_type == "efficiemcy", "efficiency")) %>%
  mutate(kitchen_type = replace(kitchen_type, kitchen_type == "efficiency ktchen", "efficiency")) %>%
  mutate(kitchen_type = replace(kitchen_type, kitchen_type == "Combo", "combo")) %>%
  mutate(kitchen_type = replace(kitchen_type, kitchen_type == "1955", NA)) 

# garage_exists column
housing_data_tbl %<>%
  mutate(garage_exists = replace(garage_exists, garage_exists == "Underground", "underground")) %>%
  mutate(garage_exists = replace(garage_exists, garage_exists == "UG", "underground")) %>%
  mutate(garage_exists = replace(garage_exists, garage_exists == "Yes", "yes")) %>%
  mutate(garage_exists = replace(garage_exists, garage_exists == "eys", "yes")) %>%
  mutate(garage_exists = replace(garage_exists, garage_exists == "1", NA))

# view cleaned housing_data_tbl
skimr::skim(housing_data_tbl)
```
# Data Manipulation (Featurization)
```{r}
# combine num_full_bathrooms and num_half_bathrooms for a new variable of num_bathroom
housing_data_tbl %<>%
  mutate(num_half_bathrooms = replace(num_half_bathrooms, is.na(num_half_bathrooms), 0)) %>%
  mutate(num_total_bathroom = num_full_bathrooms + (num_half_bathrooms / 2))

# drop columns of num_full_baths and num_half_bathrooms
housing_data_tbl %<>%
  select(-num_full_bathrooms, -num_half_bathrooms)

# multiply listing_price_to_nearest_1000 by 1000 so that sale price share 
#the same unit of dollar
housing_data_tbl %<>%
  mutate(listing_price = listing_price_to_nearest_1000 * 1000) %>%
  select(-listing_price_to_nearest_1000)

# age of apartment
housing_data_tbl$age_of_apt = 2017 - housing_data_tbl$approx_year_built

# dropped approx_year_built and garage_exists
housing_data_tbl %<>%
  select(-approx_year_built, -garage_exists) 

# combining community district values where there are less than 5 counts and labeling them other
community_district_num_counts = table(housing_data_tbl$community_district_num)
rare_levels = names(community_district_num_counts[community_district_num_counts < 20])
housing_data_tbl$community_district_num = as.character(housing_data_tbl$community_district_num)
housing_data_tbl$community_district_num[housing_data_tbl$community_district_num 
                                        %in% rare_levels] = "Other"
housing_data_tbl$community_district_num = as.factor(housing_data_tbl$community_district_num)

# combining zip codes where there are less than 5 counts and labeling them other
zip_code_counts = table(housing_data_tbl$zip_code)
rare_levels = names(zip_code_counts[zip_code_counts < 25])
housing_data_tbl$zip_code = as.character(housing_data_tbl$zip_code)
housing_data_tbl$zip_code[housing_data_tbl$zip_code %in% rare_levels] = "Other"
housing_data_tbl$zip_code = as.factor(housing_data_tbl$zip_code)

# collapsing levels of dining, dining area, none in dining room type
housing_data_tbl %<>%
  mutate(dining_room_type = replace(dining_room_type, 
                                    dining_room_type == "dining", "other")) %>%
  mutate(dining_room_type = replace(dining_room_type, 
                                    dining_room_type == "dining area", 
                                    "other")) %>%
  mutate(dining_room_type = replace(dining_room_type, 
                                    dining_room_type == "none", "other"))
housing_data_tbl$dining_room_type = as.factor(housing_data_tbl$dining_room_type)
housing_data_tbl$dining_room_type = droplevels(housing_data_tbl$dining_room_type)

# view housing_data_tbl
skimr::skim(housing_data_tbl)
```


# Joining Tables
```{r}
#left join nyc_data_per_zipcode_tbl to housing data table before casting 
#zip_code feature as a factor type
nyc_data_per_zipcode_tbl$zip_code = as.character(nyc_data_per_zipcode_tbl$zip_code)
housing_data_tbl = left_join(housing_data_tbl, nyc_data_per_zipcode_tbl, 
                             by = join_by("zip_code" == "zip_code"))

# factorize character-type features
housing_data_tbl %<>%
  mutate(across(c(cats_allowed, dogs_allowed, coop_condo, fuel_type, 
                  kitchen_type, zip_code), as.factor)) 

# view housing_data_tbl
skimr::skim(housing_data_tbl)
```

# Data Cleaning II (Missingness)
```{r}
# constructing a missingness table with 20 columns that represents 20 columns 
#with complete rate less than 1 in the original tibble table
M = as_tibble(apply(is.na(housing_data_tbl), 2, as.numeric))
colnames(M) = paste("is_missing_", colnames(housing_data_tbl), sep = "")
M %<>% 
  select_if(function(x){sum(x) > 0})

# create the design matrix by cbinding the M table - missingness itself is 
#informative. We want to preserve it in the design matrix
housing_data_miss = cbind(housing_data_tbl, M)

# smartly impute dataset using "missForest"
housing_data_imp = missForest(data.frame(housing_data_miss))$ximp

# dummify all factor variables and return a design matrix ready to be split
housing_data_imp = model.matrix(~ 0 + ., housing_data_imp)
housing_data_imp = as.data.frame(housing_data_imp)
skimr::skim(housing_data_imp)
```

# Design Matrix & Test Split
```{r}
set.seed(1717)
K = 5
# we will now work only on those rows that have an actual sale_price for training and testing
# table(housing_data_imp$is_missing_sale_price) to see for ourselves that "is_missing_sale_price == 0" 
# represents 528 rows that have the actual real world values in the outcome variable columns 
housing_design_matrix = housing_data_imp[housing_data_imp$is_missing_sale_price == 0, ]
housing_design_matrix_y = housing_design_matrix$sale_price
housing_design_matrix = as_tibble(housing_design_matrix)
housing_design_matrix %<>%
  select(-sale_price, -is_missing_sale_price)

# view housing_design_matrix
skimr::skim(housing_design_matrix)

# store indices of train set and test set
test_idx = sample(nrow(housing_design_matrix)/K, replace = FALSE)
train_idx = setdiff(1 : nrow(housing_design_matrix), test_idx)

# subsetting the training set from main model matrix
housing_data_imp_train_X = housing_design_matrix[train_idx, ]
housing_data_imp_train_y = housing_design_matrix_y[train_idx]

# subsetting the testing set from the main model matrix
housing_data_imp_test_X = housing_design_matrix[test_idx, ]
housing_data_imp_test_y = housing_design_matrix_y[test_idx]

housing_data_imp_train_X = as.data.frame(housing_data_imp_train_X)

# export housing data feature summary
housing_summary = skimr::skim(housing_data_imp)
housing_summary_df = as.data.frame(housing_summary)
write.csv(housing_summary_df, file = "housing_summary.csv", row.names = FALSE)
```

# Linear Modeling
```{r}
# vanilla OLS regression
ols_housing_model = lm(housing_data_imp_train_y ~ ., data = housing_data_imp_train_X)

# in sample metrics (from scratch)
insample_ols_hats = ols_housing_model$fitted.values
insample_ols_residuals = housing_data_imp_train_y - insample_ols_hats
insample_ols_SSE = sum(insample_ols_residuals^2)
insample_ols_SST = sum((housing_data_imp_train_y - mean(housing_data_imp_train_y))^2)
insample_ols_rsquared_from_scratch = 1 - insample_ols_SSE / insample_ols_SST # R-squared (in-sample)

insample_ols_rmse_from_scratch = sqrt( insample_ols_SSE / (nrow(housing_data_imp_train_X) - 2 ))

#in sample metrics (built-in R)
insample_ols_rsquared = summary(ols_housing_model)$r.squared # R-squared (in-sample)
insample_ols_RMSE = summary(ols_housing_model)$sigma # rmse (in-sample)

# predict on test set 
ols_y_hat = predict(ols_housing_model, newdata = housing_data_imp_test_X)

#out of sample metrics
oos_ols_residuals = housing_data_imp_test_y - ols_y_hat
SSE = sum(oos_ols_residuals^2)
SST = sum((housing_data_imp_test_y - mean(housing_data_imp_test_y))^2)

# R-squared
oos_ols_rsquared = 1 - SSE / SST

# oos RMSE
oos_ols_RMSE = sqrt(SSE / (nrow(housing_data_imp_test_X) - 2))

# show results
PercentDiff = (ols_y_hat - housing_data_imp_test_y) / housing_data_imp_test_y
PredictSummary = cbind(ols_y_hat,housing_data_imp_test_y,PercentDiff)
write.csv(PredictSummary, file = "PredictSummary.csv", row.names = FALSE)
```

# Regression Trees
```{r}
# locate the optimal node size hyperparameter for the regression tree model
max_nodesize = 70
in_sample_errors = array(NA, max_nodesize)
oos_errors = array(NA, max_nodesize)

for (i in 1 : max_nodesize){
  regression_tree_mode = YARFCART(data.frame(x = housing_data_imp_train_X), housing_data_imp_train_y, nodesize = i, calculate_oob_error = FALSE)
  y_hat = predict(regression_tree_mode, 
                  data.frame(x = housing_data_imp_train_X))
  in_sample_errors[i] = sd(housing_data_imp_train_y - y_hat)
  y_hat_test = predict(regression_tree_mode, 
                       data.frame(x = housing_data_imp_test_X))
  oos_errors[i] = sd(housing_data_imp_test_y - y_hat_test)
}

# visualize hyperparameter tunning 
ggplot(
  rbind(
    data.frame(s_e = in_sample_errors, 
               num_nodesize = 1 : max_nodesize, series = "in-sample"),
    data.frame(s_e = oos_errors, 
               num_nodesize = 1 : max_nodesize, series = "out-of-sample")
  )) +
  geom_line(aes(x = num_nodesize, y = s_e, col = series))

# optimal nodesize
optimal_nodesize = which.min(oos_errors)

# fit regression tree model using the optimal nodesize
tree_model = YARFCART(housing_data_imp_train_X, housing_data_imp_train_y, 
                      nodesize = optimal_nodesize, calculate_oob_error = TRUE, 
                      verbose = FALSE)
illustrate_trees(tree_model, max_depth = 6, open_file = TRUE)

# yhats from regression tree model
tree_model_train_yHats = predict(tree_model,housing_data_imp_train_X) 
tree_model_yHats = predict(tree_model,housing_data_imp_test_X)


# in sample R-squared
insample_regtree_residuals = housing_data_imp_train_y - tree_model_train_yHats
insample_regtree_SSE = sum(insample_regtree_residuals^2)
insample_regtree_SST = sum((housing_data_imp_train_y - mean(housing_data_imp_train_y))^2)
insample_regtree_rsquared = 1 - insample_regtree_SSE / insample_regtree_SST

# oos sample R-squared
oos_regtree_residuals = housing_data_imp_test_y - tree_model_yHats
oos_regtree_SSE = sum(oos_regtree_residuals^2)
oos_regtree_SST = sum((housing_data_imp_test_y - mean(housing_data_imp_test_y))^2)
oos_regtree_rsquared = 1 - oos_regtree_SSE / oos_regtree_SST

# RMSE
insample_rmse_regtree = sqrt(sum((housing_data_imp_train_y-tree_model_train_yHats)^2)/length(housing_data_imp_train_y)) # insample rmse

oos_rmse_regtree = sqrt(sum((housing_data_imp_test_y-tree_model_yHats)^2)/length(housing_data_imp_test_y)) # oos rmse
  
# show results
PercentDiffTree = (tree_model_train_yHats - housing_data_imp_test_y) / housing_data_imp_test_y
PredictSummaryTree = cbind(tree_model_train_yHats,housing_data_imp_test_y,PercentDiffTree)
write.csv(PredictSummaryTree, file = "PredictSummaryTree.csv", row.names = FALSE)
```

# Random Forest
```{r}
# find the best tunning parameters, starting with num_tree
num_trees_values = c(1, 2, 5, 10, 20, 30, 40, 50, 100, 200, 300, 400, 500)
oob_se_rf_mod_by_num_trees = array(NA, length(num_trees_values))
for (m in 1 : length(num_trees_values)){
  mod_rf = YARF(housing_design_matrix, housing_design_matrix_y, 
                nodesize = round(.1 * nrow(housing_data_imp_train_X)), 
                num_trees = num_trees_values[m], mtry = 60)
  oob_se_rf_mod_by_num_trees[m] = mod_rf$rmse_oob
}
# visualize hyperparameter tuning results
ggplot(data.frame(num_trees = num_trees_values, 
                  oob_se = oob_se_rf_mod_by_num_trees)) +
  aes(x = num_trees, y = oob_se) +
  geom_point() + geom_line() + scale_x_log10()

# choose optimal number of trees
optimal_num_trees = which.min(oob_se_rf_mod_by_num_trees)

# NEXT, find the best M_try parameter using the optimal num_trees
oob_se_by_mtry = array(NA, ncol(housing_data_imp_train_X))
for(m in 1: ncol(housing_data_imp_train_X)){
  mod_rf = YARF(housing_design_matrix, housing_design_matrix_y, 
                nodesize = round(.1 * nrow(housing_data_imp_train_X)), 
                num_trees = optimal_num_trees, mtry = m)
  oob_se_by_mtry[m] = mod_rf$rmse_oob
}

# visualize hyperparameter tunning results
ggplot(data.frame(num_mtry = seq(1, ncol(housing_data_imp_train_X)), 
                  oob_se = oob_se_by_mtry)) +
  aes(x = num_mtry, y = oob_se) +
  geom_point() + geom_line() 

# choose optimal mtry parameter
optimal_num_mtry = which.min(oob_se_by_mtry)

# locate the optimal node size hyperparameter for the regression tree model
max_nodesize = 70
oos_errors = array(NA, max_nodesize)
for (i in 1 : max_nodesize){
  rf_mod = YARF(housing_design_matrix, housing_design_matrix_y, nodesize = i, num_trees = optimal_num_trees, mtry = optimal_num_mtry)
  y_hat = predict(rf_mod, housing_design_matrix)
  oos_errors[i] = rf_mod$rmse_oob
}

# visualize hyperparameter tuning results
ggplot(
  rbind(
    data.frame(s_e = oos_errors, 
               num_nodesize = 1 : max_nodesize, series = "out-of-sample")
  )) +
  geom_line(aes(x = num_nodesize, y = s_e, col = series))

# choose optimal value for nodesize parameter
optimal_nodesize = which.min(oos_errors)

# let's fit a randomForest to our important feature set
rf_model = YARF(housing_data_imp_train_X, housing_data_imp_train_y, 
                nodesize = optimal_nodesize, num_trees = optimal_num_trees, 
                mtry = optimal_num_mtry, verbose = FALSE)

# out-of-bag metric
oob_rf_rmse = rf_model$rmse_oob
illustrate_trees(rf_model, max_depth = 6, open_file = TRUE)

# yhats from random forest model
rf_model_train_yHats = predict(rf_model,housing_data_imp_train_X) 
rf_model_yHats = predict(rf_model,housing_data_imp_test_X)


# in sample R-squared
insample_rf_residuals = housing_data_imp_train_y - rf_model_train_yHats
insample_rf_SSE = sum(insample_rf_residuals^2)
insample_rf_SST = sum((housing_data_imp_train_y - mean(housing_data_imp_train_y))^2)
insample_rf_rsquared = 1 - insample_rf_SSE / insample_rf_SST

# oos sample R-squared
oos_rf_residuals = housing_data_imp_test_y - rf_model_yHats
oos_rf_SSE = sum(oos_rf_residuals^2)
oos_rf_SST = sum((housing_data_imp_test_y - mean(housing_data_imp_test_y))^2)
oos_rf_rsquared = 1 - oos_rf_SSE / oos_rf_SST

# RMSE
insample_rmse_rf = sqrt(sum((housing_data_imp_train_y-rf_model_train_yHats)^2)/length(housing_data_imp_train_y)) # insample rmse

oos_rmse_rf = sqrt(sum((housing_data_imp_test_y-rf_model_yHats)^2)/length(housing_data_imp_test_y)) # oos rmse
  


# show results
print("percentage gain / loss in predictive performance")
PercentDiffTree = (tree_model_train_yHats - housing_data_imp_test_y) 
                  / housing_data_imp_test_y * 100
PredictSummaryTree = cbind(tree_model_train_yHats,housing_data_imp_test_y, 
                           PercentDiffTree)
write.csv(PredictSummaryTree, file = "PredictSummaryTree.csv", row.names = FALSE)
```

# K-fold CV 
```{r}
set.seed(1984)
n_sub = nrow(housing_design_matrix)
temp = rnorm(n_sub)

# create K equally sized folds
folds_vec = cut(temp, breaks = quantile(temp, seq(0, 1, length.out = K + 1)), 
                include.lowest = TRUE, labels = FALSE)
rm(temp)

# initialize vectors to store results
ols_rmse = numeric(K)
cart_rmse = numeric(K)
rf_rmse = numeric(K)

# define Root Mean Squared Error (RMSE) function
rmse = function(actual, predicted) {
  sqrt(mean((actual - predicted)^2))
}

# perform k-fold cross-validation
for (i in 1:K) {
  # split the data into training and testing sets
  test_indices_k = which(folds_vec == K)
  train_indices_k = which(folds_vec != K)
  
  train_data_X = housing_design_matrix[train_indices_k, ]
  train_data_y = housing_design_matrix_y[train_indices_k]
  
  test_data_X = housing_design_matrix[test_indices_k, ]
  test_data_y = housing_design_matrix_y[train_indices_k]
  
  # OLS Model
  ols_model = lm(train_data_y ~ ., data = train_data_X)
  ols_predictions = predict(ols_model, test_data_X)
  ols_rmse[i] = rmse(test_data_y, ols_predictions)
  
  # Regression Tree Model
  cart_model = YARFCART(train_data_X, train_data_y, nodesize = optimal_nodesize, calculate_oob_error = TRUE, verbose = FALSE)
  cart_predictions = predict(cart_model, test_data_X)
  cart_rmse[i] = rmse(test_data_y, cart_predictions)
  
  # Random Forest Model
  rf_model = YARF(train_data_X, train_data_y, nodesize = optimal_nodesize, 
                  num_trees = optimal_num_trees, mtry = optimal_num_mtry,
                  verbose = FALSE)
  rf_predictions = predict(rf_model, test_data_X)
  rf_rmse[i] = rmse(test_data_y, rf_predictions)
}

# calculate the mean RMSE for each of the three models
mean_ols_rmse = mean(ols_rmse)
mean_cart_rmse = mean(cart_rmse)
mean_rf_rmse = mean(rf_rmse)

# print the results
cat("Mean out-of-sample RMSE for OLS:", mean_ols_rmse, "\n")
cat("Mean out-of-sample RMSE for Regression Tree:", mean_cart_rmse, "\n")
cat("Mean out-of-sample RMSE for Random Forest:", mean_rf_rmse, "\n")
```


# Performance Validation Summary 
```{r}
# prevent table of values to express numbers in scientific notation
options(scipen=9)

# create a dataframe to report goodness of fit metrics in one clear table
goodness_of_fit_metrics = data.frame(
  error_metrics = c("in-sample R^2", "oos R^2", "in-sample RMSE", "oos RMSE", 
                    "k-fold oos RMSE"),
  OLS = c(insample_ols_rsquared, oos_ols_rsquared, insample_ols_RMSE, 
          oos_ols_RMSE, mean_ols_rmse),
  Trees = c(insample_regtree_rsquared, oos_regtree_rsquared, 
            insample_rmse_regtree, oos_rmse_regtree, mean_cart_rmse), 
  RF = c(insample_rf_rsquared, oos_rf_rsquared, insample_rmse_rf, oos_rmse_rf, 
         mean_rf_rmse )
)

colnames(goodness_of_fit_metrics) = c("Error Metrics", "OLS", 
                                      "Regression Trees", "Random Forest")

# load gt package
library(gt)

# use gt package to create a visually appealing table
final_metrics_table_gt = gt(goodness_of_fit_metrics)

final_metrics_table_gt <- 
  final_metrics_table_gt |>
  tab_header(
    title = "Goodness-of-fit Metrics"
  ) |>
  tab_spanner(
    label = "ML Models",
    columns = c("OLS", "Regression Trees", "Random Forest")
  ) |>
  cols_align(
    align = "center",
    columns = everything()
  )
  

# show the gt Goodness-of-Fit Metrics
final_metrics_table_gt

```

# Shipping Final Model Trained on All Data 
```{r}
mod_star = YARF(data.frame(housing_data_imp_train_X, housing_data_imp_test_X), data.frame(housing_data_imp_train_y, housing_data_imp_test_y), nodesize = round(.1 * nrow(housing_data_imp_train_X)), num_trees = 1000, mtry = 60, verbose = FALSE)

# ship g_final
g_final = function(x_star){
  predict(mod_star, x_star)
}
```

