---
title: "Lab 4 MATH 342W"
author: "Bryan Nevarez"
output: pdf_document
date: "11:59PM March 3rd"
editor_options: 
  markdown: 
    wrap: 72
---

Create a dataset D which we call `Xy` such that the linear model has
R\^2 about 0% but x, y are clearly associated.

```{r}
x = seq(0,6*pi, length.out = 1000)
y = sin(x)

pacman::p_load(ggplot2)
#first check that Rsq is around zero
summary(lm(y ~ x))$r.squared
#now check association visually
ggplot(data.frame(x = x, y = y)) + geom_point(aes(x = x, y = y))
```

Write a function `my_ols` that takes in `X`, a matrix with with p
columns representing the feature measurements for each of the n units, a
vector of n responses `y` and returns a list that contains the `b`, the
p+1-sized column vector of OLS coefficients, `yhat` (the vector of n
predictions), `e` (the vector of n residuals), `df` for degrees of
freedom of the model, `SSE`, `SST`, `MSE`, `RMSE` and `Rsq` (for the
R-squared metric). Internally, you cannot use `lm` or any other package;
it must be done manually. You should throw errors if the inputs are
non-numeric or not the same length. Or if `X` is not otherwise suitable.
You should also name the class of the return value `my_ols` by using the
`class` function as a setter. No need to create ROxygen documentation
here.

```{r}
my_ols = function(X, y){
  X = cbind(1,X)
  b = solve(t(X) %*% X) %*% t(X) %*% y
  
  y_hat = X %*% b
  
  e = y - y_hat 
  
  SSE = sum(e^2)
  
  SST = sum((y - mean(y))^2)
  
  df = ncol(X)
  n = nrow(X)
  MSE = SSE / (n - df)
  
  RMSE = sqrt(MSE)
  Rsq = 1 - SSE / SST    
  
  lmobj = list(
    b = b, 
    y_hat = y_hat, 
    e = e, 
    SSE = SSE, 
    SST = SST, 
    df = df, 
    MSE = MSE, 
    RMSE = RMSE, 
    Rsq = Rsq
  )
  class(lmobj) = "my_ols"
  lmobj
}
```

Verify that the OLS coefficients for the `Type` of cars in the cars
dataset gives you the same results as we did in class (i.e. the ybar's
within group).

```{r}
cars=MASS::Cars93
X= model.matrix(~Type,cars)
head(X)
y=cars$Price
my_ols(X[,-1],y)$b
```

Create a prediction method `g` that takes in a vector `x_star` and the
dataset D i.e. `X` and `y` and returns the OLS predictions. Let `X` be a
matrix with with p columns representing the feature measurements for
each of the n units

```{r}
g = function(x_star, X, y){
  c(1, x_star) = X %*% my_ols(X,y)$b
}
```

Load up the famous iris dataset. We are going to do a different
prediction problem. Imagine the only input x is Species and you are
trying to predict y which is Petal.Length. A reasonable prediction is
the average petal length within each Species. Prove that this is the OLS
model by fitting an appropriate `lm` and then using the predict function
to verify.

```{r}
data(iris)

coef(lm(Petal.Length ~ Species, iris))

mean(iris$Petal.Length[iris$Species == "setosa"])
mean(iris$Petal.Length[iris$Species == "versicolor"])
mean(iris$Petal.Length[iris$Species == "virginica"])

predict(lm(Petal.Length ~ Species, iris), data.frame(Species = "setosa"))
predict(lm(Petal.Length ~ Species, iris), data.frame(Species = "versicolor"))
predict(lm(Petal.Length ~ Species, iris), data.frame(Species = "virginica"))


```

Construct the design matrix with an intercept, X without using
`model.matrix`.

```{r}

cbind(1, ifelse(iris$Species == "versicolor", 1, 0), ifelse(iris$Species == "virginica", 1, 0))

```

We now load the diamonds dataset. Skim the dataset using skim or
summary. What is the datatype of the color feature?

ANS: The data type of the color feature is an ordered factor.

```{r}
rm(list = ls())
pacman::p_load(ggplot2, skimr)
diamonds = ggplot2::diamonds
skim(diamonds)
summary(diamonds)
class(diamonds$color)
```

Find the levels of the color feature.

```{r}
levels(diamonds$color)
```

Create new feature in the diamonds dataset, `color_as_numeric`, which is
color expressed as a continuous interval value.

```{r}
diamonds$color_as_numeric = as.numeric(diamonds$color)
head(diamonds)
```

Use that converted feature as the one predictor in a regression. How
well does this regression do as measured by RMSE?

ANS: The regression based upon this one converted feature produces a
larger RMSE as shown in the output below, and thus poor performance.

```{r}
lm(price ~ color_as_numeric, diamonds)
summary(lm(price ~ color_as_numeric, diamonds))$sigma
```

Create new feature in the diamonds dataset, `color_as_nominal`, which is
color expressed as a nominal categorical variable.

```{r}
diamonds$color_as_nominal = factor(diamonds$color, ordered = FALSE)
head(diamonds)
```

Use that converted feature as the one predictor in a regression. How
well does this regression do as measured by RMSE?

```{r}
lm(price ~ color_as_nominal, diamonds)
summary(lm(price ~ color_as_nominal, diamonds))$sigma
```

Which regression does better - `color_as_numeric` or `color_as_nominal`?
Why?

ANS: 'color_as_nominal' does better because there are more dimensions to
project upon rather than 'color_as_numeric' which is only
one-dimensional.

Now regress both `color_as_numeric` and `color_as_nominal` in a
regression. Does this regression do any better (as gauged by RMSE) than
either color_as_numeric`or`color_as_nominal\` alone?

ANS: The regression on both these features seem to produce the same rmse
as the regression upon color_as_nominal only.

```{r}
color_mod = lm(price ~ color_as_numeric + color_as_nominal, diamonds)
summary(color_mod)$sigma
```

What are the coefficients (the b vector)?

```{r}
coef(color_mod)
```

Something appears to be anomalous in the coefficients. What is it? Why?

ANS: The anomaly in the coefficients is the very last coefficient which
is NA. This is because lm is automatically incorporating the intercept
coefficient to be computed. And I also suspected that lm has a
limitation in regressing upon two different features that contain two
different datatypes.

Return to the iris dataset. Find the hat matrix H for this regression.

Find the hat matrix H for this regression of diamond price on diamond
color. Use only the first 1,000 observation in the diamond dataset.

```{r}
rm(list = ls())

diamonds1000 = ggplot2::diamonds[1:1000, ]
diamonds1000$color = factor(diamonds1000$color, ordered = FALSE)
X = model.matrix(price ~ color, diamonds1000)
H = X %*% solve(t(X) %*% X) %*% t(X)
```

Verify this hat matrix is symmetric using the `expect_equal` function in
the package `testthat`.

```{r}
pacman::p_load(testthat)
expect_equal(H,t(H)) 
```

Verify this hat matrix is idempotent using the `expect_equal` function
in the package `testthat`.

```{r}
pacman::p_load(testthat)
expect_equal(H %*% H, H)
```

Using the `diag` function, find the trace of the hat matrix.

```{r}
sum(diag(H))
Matrix::rankMatrix(H)[1]
```

It turns out the trace of a hat matrix is the same as its rank! But we
don't have time to prove these interesting and useful facts..

For masters students: create a matrix X-perpendicular.

```{r}
I = diag(nrow(H))
X_perp = I - H

y = t(t(diamonds1000$price))

X_perp %*% y

```

Using the hat matrix, compute the yhat vector and using the projection
onto the residual space, compute the e vector and verify they are
orthogonal to each other.

```{r}
y = t(t(diamonds1000$price))
yhat = H%*%y

I = diag(nrow(H))

e = (I-H) %*% y

# expect_equal(y, yhat + e)

y - (yhat + e)

t(e) %*% yhat
```

Compute SST, SSR and SSE and R\^2 and then show that SST = SSR + SSE.

```{r}
SST = sum((y - mean(y))^2)
  
SSR = sum((yhat - mean(y))^2)
  
SSE = sum(e^2)

SST - (SSR + SSE)

Rsq = SSR / SST
Rsq
```

Find the angle theta between y - ybar 1 and yhat - ybar 1 and then
verify that its cosine squared is the same as the R\^2 from the previous
problem.

```{r}
numer = sqrt(sum((yhat - mean(y))^2))
denom = sqrt(sum((y - mean(y))^2))
theta = acos(numer/denom)
cos(theta)^2
```

Project the y vector onto each column of the X matrix and test if the
sum of these projections is the same as yhat.

```{r}
n <- nrow(X)
sum_proj_y <- matrix(0,nrow = n, ncol = 1)

for(j in 1:ncol(X)){
  x_j <- X[, j, drop = FALSE]
  sum_proj_y <- sum_proj_y + (x_j %*% t(x_j) / sum(x_j^2)) %*% y
}
cbind(sum_proj_y, yhat)
#expect_equal(yhat,sum_proj_y)
sum(abs(yhat - sum_proj_y))
```

Convert this design matrix into Q, an orthonormal matrix.

```{r}
Q = matrix(NA, nrow = nrow(X), ncol = ncol(X))

Q[, 1] = X[, 1]

for(j in 2:ncol(X)){
  Q[, j] = X[, j]
  
  for(k in 1:(j-1)){
    q_k = Q[, k, drop = FALSE]
    Q[, j] = Q[, j] - ( q_k %*% t(q_k) / sum(q_k^2) ) %*% X[, j]
  }
}

for(i in 1:ncol(X)){
  Q[,i] = Q[, i] / (sqrt(sum(Q[, i])^2))
}

```

Project the y vector onto each column of the Q matrix and test if the
sum of these projections is the same as yhat.

```{r}
n <- nrow(X)
sum_proj_y <- matrix(0,nrow = n, ncol = 1)

for(j in 1:ncol(Q)){
  q_j <- Q[, j, drop = FALSE]
  sum_proj_y <- sum_proj_y + ( q_j %*% t(q_j) / sum(q_j^2) ) %*% y
}

#expect_equal(yhat, sum_proj_y)

yhat - sum_proj_y

```

Find the p=3 linear OLS estimates if Q is used as the design matrix
using the `lm` method. Is the OLS solution the same as the OLS solution
for X?

```{r}
mod_vanilla = lm(y ~ 0 + X)
b = coef(mod_vanilla)
b
mod_ortho = lm(y~0 + Q)
b_q = coef(mod_ortho)
b_q
```

Use the predict function and ensure that the predicted values are the
same for both linear models: the one created with X as its design matrix
and the one created with Q as its design matrix.

```{r}
y_hat_vanilla = predict(mod_vanilla, data.frame(X))
y_hat_ortho = predict(mod_ortho, data.frame(Q))
y_hat_vanilla
y_hat_ortho
sum(abs(y_hat_vanilla - y_hat_ortho))
```

Clear the workspace and load the boston housing data and extract X and
y. The dimensions are n = 506 and p = 13. Create a matrix that is (p +
1) x (p + 1) full of NA's. Label the columns the same columns as X. Do
not label the rows. For the first row, find the OLS estimate of the y
regressed on the first column only and put that in the first entry. For
the second row, find the OLS estimates of the y regressed on the first
and second columns of X only and put them in the first and second
entries. For the third row, find the OLS estimates of the y regressed on
the first, second and third columns of X only and put them in the first,
second and third entries, etc. For the last row, fill it with the full
OLS estimates.

```{r}
rm(list = ls()) # clear workspace

X = as.matrix(MASS::Boston[, 1 : 13]) # load Boston housing data and extract X
y = MASS::Boston$medv # load Boston housing data and extract y

p = ncol(X) # p, the number of features, is equal to the number of columns of X

# create a (p+1) x (p+1) matrix full of NA's; it is named ols_estimates
ols_estimates = matrix(NA, nrow = p+1, ncol = p+1) 

# first column name is the "intercept" column name followed by the 13 column names of X
colnames(ols_estimates) = c("intercept", colnames(X)) 

# initialize first entry of the first row as the intercept column namely ybar (null model)
ols_estimates[1,1] = coef(lm( y ~ 1))[[1]]

# implement nested for loop to iterate through the columns of X and regress y progressively onto an additional column of X at each iteration and store the values of the OLS estimates in each row of the ols_estimates matrix
for(i in 1:p){
  for(j in 1:(i+1)){
    ols_estimates[i+1,j] = coef(lm(y ~X[, 1:i]))[[j]]
  }
}

# view the matrix of OLS estimates
ols_estimates

```

Why are the estimates changing from row to row as you add in more
predictors?

ANS: The estimates change from row to row because an additional
dimension is added to the least squares algorithm. Each row contains the
OLS estimates for a hyperplane that lives in a dimension with one
additional component than the previous one. Each step of this iterative
process can be thought of the H matrix having its rank being incremented
by one and the Hy result having an additional component to its output.

Create a vector of length p+1 and compute the R\^2 values for each of
the above models.

```{r}
# initialize a vector that will contain the R^2 for each of the above models
coefs_of_deter = c(rep(NA, times = p+1))

# enter the first entry of the vector that contains the R^2 value for the null model, g(x) = ybar, since R^2 = 1 - SSE/SST = 0 (SSE = SST by definition)
coefs_of_deter[1] = 0

# populate the vector with the R^2 values for each of the above models
for(i in 1:p){
  coefs_of_deter[i+1] = summary(lm(y ~X[, 1:i]))$r.squared
}

# view vector of R^2 values (as decimals)
coefs_of_deter

```

Is R\^2 monotonically increasing? Why?

ANS: Yes, R\^2 is monotonically increasing because as we add an
additional dimensional component of data (a.k.a. "feature") to the least
squares algorithm the model will produce a "better fit" to the data. The
SSE is being further minimized with every additional feature column of
data that comes from X.

Create a 2x2 matrix with the first column 1's and the next column iid
normals. Find the absolute value of the angle (in degrees, not radians)
between the two columns in absolute difference from 90 degrees.

```{r}
# 2x2 matrix with a first column of 1's and the next column of iid normals, i.e., N(0,1)
X = cbind(matrix(1, nrow = 2, ncol = 1), matrix(rnorm(2), nrow = 2, ncol = 1))
X

# find the absolute value of the angle in degrees (not radians) between the two columns 
theta = acos((t(X[, 1]) %*% X[, 2])/(norm(X[, 1], type = "2")*norm(X[, 2], type = "2")))*(180/pi)  

# compute the absolute difference of theta from 90 degrees
abs_diff = abs(90 - theta)
abs_diff
```

Repeat this exercise `Nsim = 1e5` times and report the average absolute
angle.

```{r}
Nsim = 1e5

abs_angles = c(rep(NA, times = Nsim))

xrows = 4

for(i in 1:Nsim){
  # 2x2 matrix with a first column of 1's and the next column of iid normals, i.e., N(0,1)
X = cbind(matrix(1, nrow = xrows, ncol = 1), matrix(rnorm(xrows), nrow = xrows, ncol = 1))
X

# find the absolute value of the angle in degrees (not radians) between the two columns 
theta = acos((t(X[, 1]) %*% X[, 2])/(norm(X[, 1], type = "2")*norm(X[, 2], type = "2")))*(180/pi)  

# compute the absolute difference of theta from 90 degrees
abs_diff = abs(90 - theta)

# save output of absolute difference of theta from 90 degrees into the vector abs_angles
abs_angles[i] = abs_diff
}

# output of average absolute value (roughly 45 degrees)
mean(abs_angles)

```

Create a n x 2 matrix with the first column 1's and the next column iid
normals. Find the absolute value of the angle (in degrees, not radians)
between the two columns. For n = 10, 50, 100, 200, 500, 1000, report the
average absolute angle over `Nsim = 1e5` simulations.

```{r}
Nsim = 1e5

# defined a function (avg_abs_diff) that reports the average absolute angle difference upon the number of components of the vectors (num_times)
avg_abs_diff <- function(num_times){

  abs_angles = c(rep(NA, times = Nsim))
  for(i in 1:Nsim){
    # 2x2 matrix with a first column of 1's and the next column of iid normals, i.e., N(0,1)
  X = cbind(matrix(1, nrow = num_times, ncol = 1), matrix(rnorm(num_times), nrow = num_times, ncol =
                                                            1))

  # find the absolute value of the angle in degrees (not radians) between the two columns 
  theta = acos((t(X[, 1]) %*% X[, 2])/(norm(X[, 1], type = "2")*norm(X[, 2], type =
                                                                       "2")))*(180/pi)  

  # compute the absolute difference of theta from 90 degrees
  abs_diff = abs(90 - theta)

  # save output of absolute difference of theta from 90 degrees into the vector abs_angles
  abs_angles[i] = abs_diff
  }
  return(mean(abs_angles))
}

# results an output of the average absolute difference of angles from 90 degrees based upon the length of the vectors 
sapply(c(2, 5, 10, 50, 100, 200, 500, 1000, 5000), avg_abs_diff)
```

What is this absolute angle difference from 90 degrees converging to?
Why does this make sense?

ANS: The absolute angle difference from 90 degrees converges to 0
degrees. This convergence makes sense because it is an exhibition of the
law of large numbers which states that the average sample mean of a
sequence of random variables approaches the expectation given a
sufficient large number of them. On average the sum of the entries in
the second column vector of X will equal 0. The way we compute the angle
between two vectors requires the dot product of the first column of X
with the second column of X which results in the sum of the entries in
the second column of X that are all iid \~ N(0,1). The denominator for
the argument of inverse cosine contains the product of the first column
which will be sqrt(n) and the norm of the vector in the second column of
X. Since the norm of the second column of X is oftentimes non-zero, then
its norm will be some positive quantity and the norm of the first column
of X gets larger and larger as n increases. Thus, overall, the inverse
cosine is taking an argument that is increasingly approaching 0 which in
turn produces an output closer to 90 which implies that on average as n
increases the mean absolute angle difference converges to 0.
